---
title: "IX. Large Downloads (Experimental)"
author: "Lee Marsh"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{09. Large Downloads (Experimental)}
  %\usepackage[UTF-8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rnpn)
```

# Large downloads

`npn_export_status_data()` is an experimental alternative to
`npn_download_status_data()`. It talks to a newer service that builds your
download in the background instead of streaming it to you as it goes, which
makes it much better at large requests --- a full year of every species is
about twenty seconds of work for the server.

It is a prototype. It sits alongside the existing `npn_download_*()` functions
rather than replacing them, and its arguments may change.

```{r eval=FALSE}
obs <- npn_export_status_data(
  start_date  = "2023-01-01",
  end_date    = "2023-12-31",
  species_ids = c(3, 6, 12, 61, 82, 100)
)
```

## Dates are a span, not a list of years

There is no `years` argument. Give the span you actually want:

```{r eval=FALSE}
# spring 2020, every species
npn_export_status_data(start_date = "2020-03-01", end_date = "2020-05-31")
```

Both bounds are optional, and both accept `Date` objects or `"YYYY-MM-DD"`
strings. Omitting them asks for everything, which is a very large request.

## Choosing what you get back

The `as` argument, and nothing else, decides what comes back.

| `as` | you get |
|------|---------|
| `"data"` (default) | a tibble |
| `"path"` | the path to the downloaded `.csv.gz` file |
| `"lazy"` | a table you can query with dplyr without loading it |

Nothing else changes the return type --- not how much data matched, not
whether the download was already cached. A call that asks for a tibble either
returns a tibble or raises an error.

### `as = "path"`

Useful when you want the file itself, or when the result is too big to want in
memory at all. Nothing is read or parsed.

```{r eval=FALSE}
f <- npn_export_status_data(
  start_date  = "2023-01-01",
  end_date    = "2023-12-31",
  species_ids = 3,
  as          = "path"
)
```

### `as = "lazy"`

A lazy table lets you use dplyr verbs on data that stays on disk. Only the
answer comes back into R, when you call `collect()`:

```{r eval=FALSE}
library(dplyr)

d <- npn_export_status_data(
  start_date  = "2023-01-01",
  end_date    = "2023-12-31",
  species_ids = c(3, 6, 12, 61, 82, 100),
  as          = "lazy"
)

# how many distinct individuals of each species are monitored in each state?
by_state <- d %>%
  group_by(state, common_name) %>%
  summarise(n_individuals = n_distinct(individual_id), .groups = "drop")

by_state %>%
  group_by(common_name) %>%
  summarise(
    states        = n(),
    avg_per_state = round(mean(n_individuals), 1),
    total         = sum(n_individuals)
  ) %>%
  arrange(desc(avg_per_state)) %>%
  collect()                      # <- nothing moves until here
```

This requires the `duckdb` and `dbplyr` packages. They are checked for before
the download is submitted, so you find out immediately rather than after a long
wait.

### Reading without duckdb

`as = "data"` uses duckdb to parse the download, because it is considerably
faster on large files. If you would rather not use it --- it is a large package,
and it loads a multi-threaded C++ library that does not always play well with
debuggers --- you can switch to a pure-R reader:

```{r eval=FALSE}
options(rnpn.engine = "readr")
```

You get the same tibble, with the same columns and the same types; it is just
slower on large files. Nothing else changes, and duckdb is then never loaded at
all. `as = "lazy"` still needs duckdb, since a lazy table *is* a database
connection.

**A lazy table cannot be saved and reopened later.** It is a live connection to
a file, not data, so `saveRDS()` produces something that reloads and then errors
with `Invalid connection`. This is how dplyr behaves against every database
backend. To keep results, either `collect()` into a tibble first or use
`as = "path"` and keep the file.

## Keeping a copy of the file

`file_path` is independent of `as`. It says *where the download is kept*, not
*what you get back*, so you can ask for both:

```{r eval=FALSE}
obs <- npn_export_status_data(
  start_date  = "2023-01-01",
  end_date    = "2023-12-31",
  species_ids = 3,
  file_path   = "~/project/status-2023.csv.gz"
)
```

The file is a gzipped CSV, so name it `.csv.gz`. Note that it contains two
extra columns, `request_id` and `product`, that are dropped from the tibble;
they identify the download rather than describing an observation.

## Very large requests

When you ask for a tibble, the number of matching records is checked first, and
you get a warning if the result is unlikely to fit comfortably in memory:

```
This query matches 19,263,424 records — about 3.2 GB as a tibble, and roughly
6.0 GB while dplyr works on it. That may exceed available memory. Consider
as = "path" to download without loading, as = "lazy" to query it in place, or
narrowing your filters.
```

It is only a warning. Your download proceeds either way, and if the check
cannot be made it is skipped silently.

## Long jobs

A call waits up to `timeout` seconds (300 by default). Running out of time is
an error, but it does not cancel anything --- the server keeps working, and the
error tells you how to come back for the result:

```
Error: Job did not finish within 300s.
i It has not failed — the server is still working on it.
i Collect it later with: npn_get_job("49e2e7a8-...")
```

You can also choose not to wait at all:

```{r eval=FALSE}
job <- npn_export_status_data(start_date = "2020-01-01", wait = FALSE)

# ...later, in the same session or a different one
obs <- npn_get_job(job)
```

Results stay available on the server for about a week.

## Repeated queries

Downloads are cached, so re-running a script does not re-download the same data
or re-queue the same job. Two calls with the same filters produce the same
result even if the filters are written in a different order.

The cache lives in the session's temporary directory and disappears when R
exits. To keep it across sessions, point it somewhere durable:

```{r eval=FALSE}
options(rnpn.cache_dir = "~/npn-cache")
```

Cached files are reused for a day. To force a fresh download, pass
`refresh = TRUE`; to empty the cache, call `npn_cache_clear()`.

## A note on empty results

A filter *value* the service does not recognize returns zero rows rather than
an error --- a misspelled phenophase category gives you an empty tibble that
looks exactly like a query which genuinely matched nothing. Many legitimate
searches do return nothing, so an empty result is not by itself evidence of a
mistake. Column types are the same either way, so code that works on a full
result also works on an empty one.
